---
title:  "Reference-Free Genomic Inference with SPLASH and OASIS"
#subtitle: "a tale of two schools"
author: "[Francois Collin](https://www.linkedin.com/in/francoisz/)"
date: "`r format(Sys.time(), '%B %d, %Y')`"
always_allow_html: yes
output:
  # for use of .tabset
  bookdown::html_document2:
  #html_document:
    code_folding: hide
    code_download: true
    toc: true
    toc_depth: 3
    # does this have an effect
    fig_caption: yes
    # this has no effect
    number_sections: yes
    # css: ['../_css/pandoc3.css', '../_css/myMargins.css']
bibliography: [../_bibFiles/_healthy_aging.bib, ../_bibFiles/_ref_free_genom_inf.bib, ../../_bibFiles/_Breiman.bib, ../../_bibFiles/_Yu.bib]
csl: ../_csl/cell-numeric.csl
link-citations: true
---

```{css sidenote, echo = FALSE}

.main-container {
    margin-left: 250px;
}
.sidenote, .marginnote { 
  float: right;
  clear: right;
  margin-right: -40%;
  width: 37%;         # best between 50% and 60%
  margin-top: 0;
  margin-bottom: 0;
  font-size: 1.1rem;
  line-height: 1.3;
  vertical-align: baseline;
  position: relative;
  }
```


<style>
@import url('https://fonts.googleapis.com/css?family=Raleway');
@import url('https://fonts.googleapis.com/css?family=Oxygen');
@import url('https://fonts.googleapis.com/css?family=Raleway:bold');
@import url('https://fonts.googleapis.com/css?family=Oxygen:bold');

.main-container {
  max-width: 1400px !important;
}

body{
  font-family: 'Oxygen', sans-serif;
  font-size: 16px;
  line-height: 24px;
}

h1,h2,h3,h4 {
  font-family: 'Raleway', sans-serif;
}

.container { width: 1400px; }

caption {
  font-size: 20px;
  caption-side: top;
  text-indent: 30px;
  background-color: lightgrey;
  color: black;
  margin-top: 5px;
}

g-table-intro h4 {
  text-indent: 0px;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      fig.width = 8,
                      fig.height = 4)
```

```{r m2a-GlobalOptions, results="hide", include=FALSE, cache=FALSE}

knitr::opts_knit$set(stop_on_error = 2L) #really make it stop
options(knitr.table.format = 'html')

options(stringsAsFactors=F)

 #knitr::dep_auto()

```
<!-- ######################################################################## -->


```{r m2a-Prelims,  include=FALSE, echo=FALSE, results='hide', message=FALSE} 

FN <- "_M2A-ref_free_genomic_inf"
if(sum(grepl(FN, list.files()))==0) stop("Check FN")

 suppressMessages(require(rmarkdown))
 suppressMessages(require(knitr))

 suppressPackageStartupMessages(require(methods))
 suppressPackageStartupMessages(require(bookdown))

 suppressPackageStartupMessages(require(data.table))
 options(datatable.fread.datatable=F)

 suppressPackageStartupMessages(require(plyr))
 suppressPackageStartupMessages(require(dplyr))
 suppressPackageStartupMessages(require(magrittr))

 # Shotcuts for knitting and rendering while in R session (Invoke interactive R from R/Scripts folder)
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste(FN,".html", sep=''))

 rr <- function(n='') rmarkdown::render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep='')) ##, output_dir='Scripts')

 bb <- function(n='') browseURL(paste(FN,".html", sep=''))

 # The usual shortcuts
 zz <- function(n='') source(paste("t", n, sep=''))


 WRKDIR <- '..'
 if(!file.exists(WRKDIR)) stop("WRKDIR ERROR", WRKDIR)

 # do once

 # Shotcuts for knitting and rendering while in R session
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste('',FN,".html", sep=''))

 rr <- function(n='') render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep=''), output_dir='Scripts')

 bb <- function(n='') browseURL(paste('',FN,".html", sep=''))

 # The usual shorcuts
 zz <- function(n='') source(paste('', "t", n, sep=''))

 # file rmarkdown file management options: cache, figures
 cache_DIR <- file.path(WRKDIR,'Scripts', 'cache/M2A/')
 suppressMessages(dir.create(cache_DIR, recursive=T))
 opts_chunk$set(cache.path=cache_DIR)

 figures_DIR <- file.path(WRKDIR,'Scripts', 'figures/M2A/')
 suppressMessages(dir.create(figures_DIR, recursive=T))
 opts_chunk$set(fig.path=figures_DIR)

 #tables_DIR <- file.path(WRKDIR,'Scripts', 'tables/M2A/')
 #suppressMessages(dir.create(table_DIR, recursive=T))
 #opts_chunk$set(fig.path=table_DIR)
 
 # need a local copy of help_DIR
 #help_DIR <- file.path(WRKDIR,'Scripts', 'help_files')
 help_DIR <- file.path('.', 'help_files')
 suppressMessages(dir.create(help_DIR, recursive=T))
 
 temp_DIR <- file.path(WRKDIR,'Scripts', 'temp_files')
 suppressMessages(dir.create(temp_DIR, recursive=T))

```
<!-- ######################################################################## -->

*** 

```{r m2a-utilityFns, echo=FALSE}
 # Here we define some utility functions
source('r/utilityFns.r')

```

<!-- ######################################################################## -->


<!-- SKIP THIS 
# Params {#params}

## Input params {-}
-->

```{r m2a-input-params-all, echo = FALSE, results = "asis", eval=F, echo=F}
 # print out original input params
data.frame(
param_name=names(params),
param_value=unlist(params),
param_class=sapply(params, class), row.names=NULL) %>%
  knitr::kable(caption="Input Parameters")
```

<br/>


<!--
* Breiman: [@Breiman:1984aa; @Breiman:2001aa]
* Freedman: [@Freedman:2008aa; @Freedman:2008ad; @Freedman:2008ac; @Freedman:2009ac]
-->

## Preliminaries {-}

* chatGPT query:
   - what is the best GPT based tool to obtain definitions of technical terms used in biotech related scientific publications?

* Response

### GPT-based tools for technical terms  {-}

* Best GPT-based tools for defining technical terms in biotech-related scientific publications:

| Tool | Strengths | Ideal Usage |
|------|-----------|-------------|
| **[SciSpace Copilot](https://scispace.com)** | - Contextual definitions within scientific papers<br>- Highly accurate for biotech terms<br>- Integrated directly into research PDFs | Immediate comprehension within biotech research papers |
| **[Elicit.org](https://elicit.org)** | - GPT-powered scientific research assistant<br>- Simplifies technical jargon into plain language<br>- Effective for quick clarification | Rapid clarification and concise definitions |
| **[Explainpaper](https://www.explainpaper.com)** | - GPT-driven academic language translator<br>- Clear, user-friendly explanations of complex biotech concepts | Students or researchers seeking straightforward definitions |
| **[Consensus](https://consensus.app)** | - Aggregates and summarizes definitions from multiple scientific papers<br>- Provides context-aware definitions within the broader biotech field | Cross-validating definitions across multiple sources |

#### Recommended overall choice:  {-}
- ✅ **[SciSpace Copilot](https://scispace.com)** is the best overall choice for its contextual accuracy and specialized biotech literature integration. 

**Alternatives:** Use **Elicit** or **Consensus** if definitions from broader perspectives or cross-validation across sources is desired.

<br/>


# Introduction

* The [Salzman Lab](https://salzmanlab.stanford.edu/) 
develop and apply new **statistical algorithms** for **biological inference**
on fundamental questions in the evolution of genomes.

<p><p/>
* In this report we focus on two papers from the 
[Salzman Lab](https://salzmanlab.stanford.edu/):
   - SPLASH: A statistical, reference-free genomic algorithm unifies biological discover
[PDF](https://www.cell.com/cell/pdf/S0092-8674(23)01179-0.pdf),
[Code](https://github.com/refresh-bio/SPLASH) [@Chaung:2023aa]
   - OASIS: An interpretable, finite-sample valid alternative to Pearson’s X2 
for scientific discovery
[URL](https://www.pnas.org/doi/10.1073/pnas.2304671121),
[Code](https://github.com/refresh-bio/SPLASH) [@Baharav:2024aa]


## Outline  {-}

<p><p/>
* In Section \@ref(GPT-based-res) we prompt **chatGPT** to compare the following
research assistant tools: (also done in [_M1A-healthy_aging_genes.html](./M1A-healthy_aging_genes.html)
   - **Scholar GPT, Concensus, SciSpace, Semantic Scholar, Perplexity, Elicit**

<p><p/>
* In Section \@ref(soa-ref-free-genomics), we use **Concensus** to obtain a
literature review summarizing the state of the art in reference-free genomics.


<p><p/>
* In Section \@ref(splash-ai-review) we prompt **chatGPT** and **SciSpace** to provide a
technical review of Chaung et al. (2023) [@Chaung:2023aa] (SPLASH).
   - **does** the AI assistants get some key concepts wrong while maintaining 
a steady tone of confidence.

<p><p/>
* In Section \@ref(oasis-ai-review) we prompt **chatGPT** and **SciSpace** to provide a
technical review of Baharav et al. (2024) [@Baharav:2024aa].
**Does** the AI assistants get some
key concepts wrong while maintaining a steady tone of confidence.

<!--
<p><p/>
* In the next Section, \@ref(oasis-ai-review-with-line-ref), we prompt
**chatGPT and **SciSpace** to provide line references for all claims or attributions
made regarding the article.
   - How much does the response change?

<p><p/>
* In the next Section, \@ref(splash-ai-review-with-line-ref), we prompt
**chatGPT and **SciSpace** to provide line references for all claims or attributions
made regarding the article.
   - How much does the response change?

-->

<br/>


## Upload articles {-}

* In this report we focus on two papers from the [Salzman Lab]:
   - SPLASH: 
A statistical, reference-free genomic algorithm unifies biological discover
[PDF](https://www.cell.com/cell/pdf/S0092-8674(23)01179-0.pdf),
[Code](https://github.com/refresh-bio/SPLASH) [@Chaung:2023aa]
[local](/Users/fcollin/Documents/Projects/reAnalysis/Refs/Ref-free-genom-inf/Chaung_2023.pdf)

   - OASIS: 
An interpretable, finite-sample valid alternative to Pearson’s X2 
for scientific discovery
[URL](https://www.pnas.org/doi/10.1073/pnas.2304671121),
[Code](https://github.com/refresh-bio/SPLASH) [@Baharav:2024aa]
[local](/Users/fcollin/Documents/Projects/reAnalysis/Refs/Ref-free-genom-inf/Baharav_2024.pdf,Baharav_2024_supp.pdf)

* Upload to project(s)


<br/>



# GPT-Based Research Assistants {#GPT-based-res .tabset}


* The results in this section are most un-impressive.
* Not sure how to make them better
)

* Before diving into the articles of interest, we will submit queries to 
establish the current state of scientific knowledge regarding 
reference-free genomics.

<p><p/>
* There are several GPTs which are specialized for research:
   - Scholar GPT
   - Concensus
   - SciSpace
   - Semantic Scholar
   - Perplexity
   - Elicit

<p><p/>
* See [Consensus GPT vs SciSpace GPT](https://www.youtube.com/watch?v=EN0JZTvbuzQ)
for a comparison of these two GPTs.

<br/>

## Home {-}

<br/>
<br/>

## chatGPT Prompt - 4 tools {-}

* Reference-free


* Compare and contrast 4 research tools - 
Scholar GPT, Concensus, SciSpace, Semantic Scholar -
in their strengths and weaknesses when used to establish
the state of the art knowledge  in reference-free genomics

<br/>


## **Response** {-}

<br/>
<br/>

## chatGPT Prompt - 6 tools {-}


* Compare and contrast 6 research tools - 
Scholar GPT, Concensus, SciSpace, Semantic Scholar, Elicit, and Perplexity -
in their strengths and weaknesses when used to establish
the state of the art knowledge  in reference-free genomics

<br/>

## **Response** {-}

<br/>
<br/>

# State of the Arts:  Reference-Free Genomic Inference {#soa-ref-free-genomics  .tabset}


* In this report we focus on two papers from the [Salzman Lab]:
   - SPLASH: 
A statistical, reference-free genomic algorithm unifies biological discover
[PDF](https://www.cell.com/cell/pdf/S0092-8674(23)01179-0.pdf),
[Code](https://github.com/refresh-bio/SPLASH) [@Chaung:2023aa]
   - OASIS: 
An interpretable, finite-sample valid alternative to Pearson’s X2 
for scientific discovery
[URL](https://www.pnas.org/doi/10.1073/pnas.2304671121),
[Code](https://github.com/refresh-bio/SPLASH) [@Baharav:2024aa]

## Home {-}

<br/>
<br/>

## Consensus Prompt {-}

* reference-free genomic inference is part of the methodologies discussed in
the two attached papers.  Please produce a literature review to report
on the state of the art knowledge of reference-free genomic inference.

* We are particularly interested in the statistical aspects of methods.

* Keep the report in un-rendered markdown format, or markdown syntax
* Also include the markdown syntax for a bibliography of citations."

<br/>
<br/>

## **Response**  {-}



<br/>
<br/>

# SPLASH - Chaung et al, 2023 [@Chaung:2023aa] {.tabset}

## Home {-}

<br/>


## chatGPT Prompt {-}

<br/>


## **Response**  {-}

<br/>



## SciSpace Prompt {-}


<br/>


## **Response**  {-}

<br/>

# OASIS -  Baharav et al., 2024 [@Baharav:2024aa] {.tabset}

## Home {-}

<br/>


## chatGPT Prompt {-}

<br/>


## **Response**  {-}

<br/>



## SciSpace Prompt {-}


<br/>


## **Response**  {-}

<br/>





# Appendix 

<br/>

## What is a Statistical Algorithm?

* Submitted to SciSpace (copilot)

A **statistical algorithm** is a computational procedure that applies statistical techniques to analyze data, make predictions, or identify patterns. These algorithms use probability theory, statistical inference, and mathematical modeling to extract meaningful insights from data. They are widely used in fields such as machine learning, data science, econometrics, and bioinformatics.

### Key Characteristics:
1. **Data-Driven**: Statistical algorithms rely on observed data to make estimations or predictions.
2. **Probabilistic Nature**: They incorporate probability distributions to model uncertainties.
3. **Optimization-Based**: Many involve optimization techniques to minimize errors and improve accuracy.
4. **Inference and Estimation**: They estimate parameters, test hypotheses, and infer relationships.

### Examples of Statistical Algorithms:
- **Regression Models**: Linear Regression, Logistic Regression
- **Classification Models**: Naïve Bayes, Decision Trees (with statistical criteria like Gini Impurity or Entropy)
- **Clustering Methods**: K-Means, Gaussian Mixture Models (GMM)
- **Dimensionality Reduction**: Principal Component Analysis (PCA), Singular Value Decomposition (SVD)
- **Bayesian Inference**: Markov Chain Monte Carlo (MCMC), Hidden Markov Models (HMM)

### Applications:
- **Machine Learning**: Feature selection, model training, and prediction.
- **Finance**: Risk assessment, stock market predictions.
- **Healthcare**: Disease diagnosis, medical imaging analysis.
- **Social Sciences**: Survey analysis, behavioral predictions.

In essence, statistical algorithms form the backbone of data-driven decision-making, allowing machines and researchers to extract patterns, make predictions, and improve decision accuracy.


## What is Biological Inference?

* Submitted to SciSpace (copilot)

**Poor definition.**

**Biological inference** refers to the process of drawing conclusions about biological systems, processes, or phenomena using data analysis, statistical models, and computational methods. It involves using observed biological data to make predictions, identify patterns, or infer underlying mechanisms.

### Key Aspects of Biological Inference:
1. **Data-Driven Analysis**: Utilizes experimental or observational biological data.
2. **Statistical and Computational Models**: Applies statistical inference, machine learning, and bioinformatics tools.
3. **Hypothesis Testing**: Evaluates scientific hypotheses based on data.
4. **Predictive Insights**: Helps predict biological behaviors and interactions.

### Examples of Biological Inference:
- **Genomic Inference**: Identifying genetic variations linked to diseases.
- **Phylogenetic Inference**: Reconstructing evolutionary relationships among species.
- **Protein Structure Prediction**: Inferring 3D structures from amino acid sequences.
- **Ecological Inference**: Understanding species distribution and population dynamics.
- **Systems Biology**: Modeling complex biological networks and interactions.

### Applications:
- **Medicine**: Predicting disease susceptibility and drug responses.
- **Evolutionary Biology**: Understanding the genetic basis of adaptation.
- **Ecology**: Assessing environmental impacts on biodiversity.
- **Bioinformatics**: Analyzing large-scale biological datasets.

biological inference plays a crucial role in advancing our understanding of life sciences 
by integrating data-driven approaches to uncover hidden patterns and 
relationships in biological systems.



<br/>
<br/>

# References {#references}
<div id="refs"></div>
    
<br/>

# Appendix {-}

## Parameter settings {-}
  * WRKDIR = `r normalizePath(WRKDIR)`
  * FN = `r FN`
  * Scripts = Scripts
  * RUN DATE = `r date()`

```{r m2a-session-info, echo=T, eval=T}
 sessionInfo()
```

```{r, echo=FALSE}
  knit_exit()
```



## chatGPT Prompt {-}

* Read the attached paper by Sood et.al. 2015 [@Sood:2015aa].

* Act as a statistician who is proficient in the traditional statistical concepts as they apply to the analysis of designed experiments, as well as in the novel statistical concepts developed for the analysis of omic data, and review the article.

* The review should not consider findings from later studies

* The review should focus on the statistical aspects of the data collected and analysed.
* Key statistical principles include:
  - verifying that the training data points can be thought of as being
a representative, random subset of the target population
  - a representative subset most include enough of the right sort of sample.
   - how many data collection runs are aggregated or pooled to assemble the analysis dataset?
   - how are the different tissue types distributed across data generation runs?  
   - Is run associated with age. 
   - Were the data from different runs normalized?
* What evidence is provided that the classifiers developed work well with datasets analyzed and will continue to work well when future data are analyzed?

* Keep the report in un-rendered markdown format, or markdown syntax.
* Also include the markdown syntax for a bibliography of citations.

<br/>

## chatGPT RE:Prompt {-}

Note that what the authors  call external validation  entails fitting a new model with each dataset.  There is no validation of a model; only validation that the subset of 150 probe sets are sufficient to enable KNN + LOOCV to give rise to a fitted model that still discriminates.  This is not validation in the usual sense.  There is no attempt to compare the fitted models from the different data sets.  Jacob and Speed (attached) show that repeating the fit and eval procedures with a random selection of 150 probe sets achieves similar discrimination on some datasets.
Please include these facts in the conclusions to be drawn form these data.

<br/>

* The review should not consider findings from later studies

* The review should focus on the statistical aspects of the data collected and analysed.
* Key statistical principles include:
   - how are the different tissue types distributed across data generation runs?  
   - Is run associated with age. 
   - Were the data from different runs normalized?
* What evidence is provided that the classifiers developed work well with datasets analyzed and will continue to work well when future data are analyzed?

* Keep the report in un-rendered markdown format, or markdown syntax.
* Also include the markdown syntax for a bibliography of citations.


<br/>
<br/>


<!-- To run
# nohup Rscript -e "knitr::knit2html('_M2A-ref_free_genomic_inf.Rmd')" > _M2A-ref_free_genomic_inf.log  &

# Or
# nohup Rscript -e "rmarkdown::render('_M2A-ref_free_genomic_inf.Rmd')" > _M2A-ref_free_genomic_inf.log  &

-->

