---
title: "Making Neyman (1923) insights accessible"
#subtitle: ""
author: "[Francois Collin](https://www.linkedin.com/in/francoisz/)"
#date: '`r format(Sys.time(), "%a %b %d", tz="America/Los_Angeles")`'
always_allow_html: yes
output:
  html_document:
    code_folding: hide
    code_download: true
    toc: true
    toc_depth: 2
    # does this have an effect
    fig_caption: yes
    # this has no effect
    number_sections: yes
    # css: ['../_css/pandoc3.css', '../_css/myMargins.css']
bibliography: [../../../_bibFiles/_Normalization.bib,../../../_bibFiles/_RUV.bib,../../Refs/rand_cause_inf/_rand_cause_inf.bib]
csl: ../../../_csl/cell-numeric.csl
link-citations: true
---

<!--
title: "Randomization Based Inference: Neyman (1923) On the Application of Prob Theory"
-->

`r DT_TABLES <- F`

```{css sidenote, echo = FALSE}

.main-container {
    margin-left: 250px;
}
.sidenote, .marginnote { 
  float: right;
  clear: right;
  margin-right: -40%;
  width: 37%;         # best between 50% and 60%
  margin-top: 0;
  margin-bottom: 0;
  font-size: 1.1rem;
  line-height: 1.3;
  vertical-align: baseline;
  position: relative;
  }
```


<style>
@import url('https://fonts.googleapis.com/css?family=Raleway');
@import url('https://fonts.googleapis.com/css?family=Oxygen');
@import url('https://fonts.googleapis.com/css?family=Raleway:bold');
@import url('https://fonts.googleapis.com/css?family=Oxygen:bold');

.main-container {
  max-width: 1400px !important;
}

body{
  font-family: 'Oxygen', sans-serif;
  font-size: 16px;
  line-height: 24px;
}

h1,h2,h3,h4 {
  font-family: 'Raleway', sans-serif;
}

.container { width: 1400px; }

caption {
  font-size: 20px;
  caption-side: top;
  text-indent: 30px;
  background-color: lightgrey;
  color: black;
  margin-top: 5px;
}

g-table-intro h4 {
  text-indent: 0px;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      fig.width = 8,
                      fig.height = 4)
```

```{r m5a-GlobalOptions, results="hide", include=FALSE, cache=FALSE}

knitr::opts_knit$set(stop_on_error = 2L) #really make it stop
options(knitr.table.format = 'html')

options(stringsAsFactors=F)

 #knitr::dep_auto()

```
<!-- ######################################################################## -->


```{r m5a-Prelims,  include=FALSE, echo=FALSE, results='hide', message=FALSE} 

FN <- "_M5A-Neyman_1923"
if(sum(grepl(FN, list.files()))==0) stop("Check FN")

 suppressMessages(require(rmarkdown))
 suppressMessages(require(knitr))

 suppressPackageStartupMessages(require(methods))
 suppressPackageStartupMessages(require(bookdown))

 suppressPackageStartupMessages(require(data.table))
 options(datatable.fread.datatable=F)

 suppressPackageStartupMessages(require(plyr))
 suppressPackageStartupMessages(require(dplyr))
 suppressPackageStartupMessages(require(magrittr))

 # Shotcuts for knitting and rendering while in R session (Invoke interactive R from R/Scripts folder)
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste(FN,".html", sep=''))

 rr <- function(n='') rmarkdown::render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep='')) ##, output_dir='Scripts')

 bb <- function(n='') browseURL(paste(FN,".html", sep=''))

 # The usual shortcuts
 zz <- function(n='') source(paste("t", n, sep=''))


 WRKDIR <- '..'
 if(!file.exists(WRKDIR)) stop("WRKDIR ERROR", WRKDIR)

 # do once

 # Shotcuts for knitting and rendering while in R session
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste('',FN,".html", sep=''))

 rr <- function(n='') render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep=''), output_dir='Scripts')

 bb <- function(n='') browseURL(paste('',FN,".html", sep=''))

 # The usual shorcuts
 zz <- function(n='') source(paste('', "t", n, sep=''))

 # file rmarkdown file management options: cache, figures
 cache_DIR <- file.path(WRKDIR,'Scripts', 'cache/M5A/')
 suppressMessages(dir.create(cache_DIR, recursive=T))
 opts_chunk$set(cache.path=cache_DIR)

 figures_DIR <- file.path(WRKDIR,'Scripts', 'figures/M5A/')
 suppressMessages(dir.create(figures_DIR, recursive=T))
 opts_chunk$set(fig.path=figures_DIR)

 #tables_DIR <- file.path(WRKDIR,'Scripts', 'tables/M5A/')
 #suppressMessages(dir.create(table_DIR, recursive=T))
 #opts_chunk$set(fig.path=table_DIR)
 
 # need a local copy of help_DIR
 #help_DIR <- file.path(WRKDIR,'Scripts', 'help_files')
 help_DIR <- file.path('.', 'help_files')
 suppressMessages(dir.create(help_DIR, recursive=T))
 
 temp_DIR <- file.path(WRKDIR,'Scripts', 'temp_files')
 suppressMessages(dir.create(temp_DIR, recursive=T))

```
<!-- ######################################################################## -->

*** 

```{r m5a-utilityFns, echo=FALSE}
 # Here we define some utility functions
source('r/utilityFns.r')

```

<!-- ######################################################################## -->


<!-- SKIP THIS 
# Params {#params}

## Input params {-}
-->

```{r m5a-input-params-all, echo = FALSE, results = "asis", eval=F, echo=F}
 # print out original input params
data.frame(
param_name=names(params),
param_value=unlist(params),
param_class=sapply(params, class), row.names=NULL) %>%
  knitr::kable(caption="Input Parameters")
```

<br/>

<!-- DNA
# Preliminaries {-}
 
## Read Support Data {-} 
# To Do

-->

# Introduction  {#intro}

* This article is a summary of 
*From urn models to box models: Making Neyman's (1923) insights accessible*
[@Lin:2024aa].

* Neyman’s 1923 paper introduced the potential outcomes framework and the 
foundations of randomization-based inference about average treatment effects. 
([@Splawa-Neyman:1990aa;@Speed:1990aa;@Rubin:1990aa]).

* In the context of an agricultural experiment, Neyman assumes that the assignment of treatments
(crop varieties) to units (plots of land) is analogous to randomly drawing balls from urns without
replacement, so that each treatment group is a simple random sample from the finite population
of experimental units.

* Although he does not propose randomization in a literal sense [2,3], his urn model is
equivalent to a completely randomized experiment, and he considers the problem of estimating
the average effect of treatment i relative to treatment j.

<p><p/>
* In his framework for statistical inference, the units in the experiment are not assumed
to be a random sample from a larger target population.
   - Instead, the experimental units are the target population, and random assignment
of treatment is the sole source of randomness.

<br/>
<br/>

# Results  

<p><p/>
* Neyman’s first key result follows immediately from the properties of simple random samples:
   - the difference in mean outcomes between treatment groups i and j is an unbiased estimator
of the average treatment effect:
      - $\widehat{ATE} = \widehat{\bar{y(1)}} -  \widehat{\bar{y(0)}}$


<p><p/>
* Consider the usual estimator of the variance of the difference-in-means estimator: 
   - $\frac{s_i^2}{n_i}+\frac{s_j^2}{n_j}$
   - A traditional justification for this formula assumes
**two independent random samples from two infinite populations**.  

<p><p/>
* Neyman’s framework violates these assumptions:
   - the population is **finite** and the treatment groups are mutually exclusive, **not independent**
   - but surprisingly, he finds that:
      -  the usual variance estimator is unbiased 
if there is a constant additive treatment effect and 
      - the usual variance estimator is conservatively biased
if there are heterogeneous treatment effects.


<br/>

# Details

## Setup

<p><p/>
* N subjects; Two disjoint simple random samples of fixed sizes 
$n_1$ and $n_0$ are assigned to Treatment and Control groups.

<p><p/>
* Constants:
   - $y_i(1)$, $y_i(0)$ denote subject $i$'s potential outcomes under 
treatment and control, resp..

<p><p/>
* Random variables:
   - $(T_i = 1)  \text{ if subject i is in Treatment group}$
$(T_i = 0)  \text{ Otherwise}$
   - $(C_i = 1)  \text{ if subject i is in Control group}$
$(C_i = 0)  \text{ Otherwise}$

<br/>

* The true Average Treatment Effect is:


\begin{equation}
\mathrm{ATE}=\bar{y}(1)-\bar{y}(0)
\end{equation}

where
\begin{equation}
\bar{y}(1)=\frac{1}{N} \sum_{i=1}^N y_i(1) \text { and } \\
\bar{y}(0)=\frac{1}{N} \sum_{i=1}^N y_i(0)
\end{equation}


* The difference-in-means estimator of ATE
is defined as follows:

\begin{equation}
\begin{aligned}
&\widehat{\mathrm{ATE}}=\widehat{\bar{y}}(1)-\widehat{\bar{y}}(0)
\end{aligned}
\end{equation}

where 
\begin{equation}
\begin{aligned}
&\widehat{\bar{y}}(1)=\frac{1}{n_1} \sum_{i=1}^N T_i y_i(1) \text { and } \\
&\widehat{\bar{y}}(0)=\frac{1}{n_0} \sum_{i=1}^N C_i y_i(0) 
\end{aligned}
\end{equation}



<br/>
<br/>

## Unbiased estimation of the average treatment effect

* The calculation  hinges on the assumption that treatment assignment is random and that the probability of assignment is uniform across units.

<br/>

### **Expectation of $T_i$ under Random Assignment** {-}

* Assuming each unit is assigned to treatment independently with probability
$\frac{n_1}{N}$.

$$E(T_i) = P(T_i = 1) = \frac{n_1}{N}$$

<br/>

### **Substitute E($T_i$) and Simplify**:   {-}

\begin{equation}
E[\widehat{\bar{y}}(1)]=\frac{1}{n_1} \sum_{i=1}^N E\left[T_i\right] y_i(1)=\frac{1}{n_1} \sum_{i=1}^N \frac{n_1}{N} y_i(1)=\bar{y}(1) .
\end{equation}


<br/>
<br/>


## True variance of the difference-in-means estimator


* See online article: https://www.degruyter.com/document/doi/10.1515/jci-2023-0073/html?srsltid=AfmBOoofY9z3KKX243onbmNVppAih3V8XeECOaaNXAGjsMpi4qQLRK7C#j_jci-2023-0073_ref_037

* The variance of the difference-in-means estimator is

\begin{equation}
\operatorname{Var}\left(\widehat{\operatorname{ATE}}\right)=\operatorname{Var}(\widehat{\bar{y}}(1))+\operatorname{Var}(\widehat{\bar{y}}(0))-2 \operatorname{Cov}(\widehat{\bar{y}}(1), \widehat{\bar{y}}(0))
\end{equation}

* Using well-known formulas for the **variance of the sample mean**
under simple random sampling without replacement

\begin{equation}
\operatorname{Var}(\widehat{\bar{y}}(1))=\left(1-\frac{n_1-1}{N-1}\right) \frac{\sigma_1^2}{n_1}=\left(1-\frac{n_1}{N}\right) \frac{\tilde{\sigma}_1^2}{n_1},
\end{equation}

where
\begin{equation}
\sigma_1^2=\frac{1}{N} \sum_{i=1}^N\left[y_i(1)-\bar{y}(1)\right]^2   \text { and } \\
\tilde{\sigma}_1^2=\frac{1}{N-1} \sum_{i=1}^N\left[y_i(1)-\bar{y}(1)\right]^2
\end{equation}

Similarly for $\operatorname{Var}(\widehat{\bar{y}}(0))$

\begin{equation}
\operatorname{Cov}(\widehat{\bar{y}}(1), \widehat{\bar{y}}(0))=-\frac{1}{N-1} \sigma_{1,0}=-\frac{1}{N} \tilde{\sigma}_{1,0}
\end{equation}

where 
\begin{equation}
\sigma_{1,0}=\frac{1}{N} \sum_{i=1}^N\left[y_i(1)-\bar{y}(1)\right]\left[y_i(0)-\bar{y}(0)\right]
    \text { and } \\
\tilde{\sigma}_{1,0}=\frac{1}{N-1} \sum_{i=1}^N\left[y_i(1)-\bar{y}(1)\right]\left[y_i(0)-\bar{y}(0)\right]
\end{equation}



For proof see online article:
[From urn models to box models: Making Neyman's (1923) insights accessible](https://www.degruyter.com/document/doi/10.1515/jci-2023-0073/html?srsltid=AfmBOoofY9z3KKX243onbmNVppAih3V8XeECOaaNXAGjsMpi4qQLRK7C#j_jci-2023-0073_ref_037)



<br/>
<br/>

* **Putting it all together**

\usepackage{amsmath}
\begin{equation}
\tag{A.1}
\operatorname{Var}\left(\widehat{\mathrm{ATE}}\right) =
\left(1-\frac{n_1}{N}\right) \frac{\tilde{\sigma}_1^2}{n_1} +
\left(1-\frac{n_0}{N}\right) \frac{\tilde{\sigma}_0^2}{n_0} +
2 \frac{\tilde{\sigma}_{1,0}}{N}
\label{eq:var-ate-hat}
\end{equation}


* **Rearranging**

\begin{equation}

\operatorname{Var}\left(\widehat{\mathrm{ATE}}\right)=\frac{\tilde{\sigma}_1^2}{n_1}+\frac{\tilde{\sigma}_0^2}{n_0}-\frac{\tilde{\sigma}_1^2+\tilde{\sigma}_0^2-2 \tilde{\sigma}_{1,0}}{N}

\end{equation}

* which simplifies to

\begin{equation}
\tag{A.2}
\operatorname{Var}\left(\widehat{\mathrm{ATE}}\right)=\frac{\tilde{\sigma}_1^2}{n_1}+\frac{\tilde{\sigma}_0^2}{n_0}-\frac{\tilde{\sigma}_{\Delta}^2}{N}

\end{equation}

where  
$$\tilde{\sigma}_{\Delta}^2=\frac{1}{N-1} \sum_{i=1}^N\left[y_i(1)-y_i(0)-\mathrm{ATE}\right]^2$$
is the population variance of the treatment effect
(thinking of the N subjects as the finite population,
and again using the denominator N − 1).

<br/>
<br/>

## Unbiased or conservatively biased estimation of the variance



* The usual estimator for $\operatorname{Var}]\left(\widehat{\mathrm{ATE}}\right)$ is:

\begin{equation}
\tag{A.3}
\widehat{\operatorname{Var}}\left(\widehat{\mathrm{ATE}}\right)=\frac{s_1^2}{n_1}+\frac{s_0^2}{n_0}
\end{equation}

where
\begin{equation}
s_1^2=\frac{1}{n_1-1} \sum_{i=1}^N T_i\left[y_i(1)-\widehat{\bar{y}}(1)\right]^2 
    \text { and } \\
s_0^2=\frac{1}{n_0-1} \sum_{i=1}^N C_i\left[y_i(0)-\widehat{\bar{y}}(0)\right]^2
\end{equation}

are the sample variances of the observed outcome in the treatment group and control group.

* From finite-population sampling theory, $s_1^2$ and $s_0^2$ are unbiased estimators of $\tilde{\sigma}_1^2$ and $\tilde{\sigma}_0^2$.  Thus,

\usepackage{amsmath}
\begin{equation}
\tag{A.4}
E[\widehat{\operatorname{Var}}(\widehat{\mathrm{ATE}})]=\frac{\tilde{\sigma}_1^2}{n_1}+\frac{\tilde{\sigma}_0^2}{n_0}
\label{eq:exp-var-hat-ate-hat}
\end{equation}


<p><p/>
* Comparing A.4 with A.1 we can see that in average,
$\widehat{\operatorname{Var}}(\widehat{\mathrm{ATE}})$
makes two mistakes: finite populations and positive correlations:
   -  The first mistake is conservative: it ignores the finite-population correction factors,
(1 - $n_1$/N) and (1 - $n_0$/N).
   - second mistake is anticonservative if the two potential outcomes are
positively correlated in the population of N subjects


<p><p/>
* Comparing A.4 with A.2, we  can see the **net effect of the two mistakes**. 
   - the net effect is nil **if there is a constant additive treatment effect**: 
$y_i(1) - y_i(0)$ = k.
   - Otherwise $\widehat{\mathrm{Var}}(\widehat{\mathrm{ATE}})$
has a conservative bias and undersattes the true variability by
$\tilde{\sigma}_{\Delta}^2 / N$.

<p><p/>
* The pooled estimator of $\operatorname{Var}(\widehat{\mathrm{ATE}})$:

\begin{equation}
\tag{A.5}
\widehat{\operatorname{Var}}(\widehat{\mathrm{ATE}})_p = 
\frac{\left(n_1-1\right) s_1^2+\left(n_0-1\right) s_0^2}{n_1+n_0-2}\left[\frac{1}{n_1}+\frac{1}{n_0}\right] .
\end{equation}


* A5 is identical to  A3 if $n_0 = n_1$ and is unbiased if
there is a constant additive treatment effect.

* If $n_0 \neq n_1$ and there are heterogeneous treatment effects,
the pooled estimator may be either conservative or anticonservative.

<br/>
<br/>

# References {#references}
<div id="refs"></div>
    
<br/> 

```{r , eval=T}
pander::pander(sessionInfo())
```

  * WRKDIR = `r normalizePath(WRKDIR)`
  * FN = `r FN`
  * Scripts = Scripts
  * RUN DATE = `r date()`

```{r , echo=FALSE}
  knit_exit()
```

##########################################################
### ARCHIVE
##########################################################

### **Definition of the Estimator**: {-}

\begin{equation}

\widehat{\bar{y(1)}} = \frac{\sum_{i=1}^N T_i y_i(1)}{n_1}

\end{equation}

<br/>

### **Taking Expectation**:  {-}

\begin{equation}

E \left[ \widehat{\bar{y(1)}} \right] = E \left[  \frac{\sum_{i=1}^N T_i y_i(1)}{n_1} \right]

\end{equation}

<br/>

### **By linearity of Expectation**   {-}

\begin{equation}

E  \left[  \frac{\sum_{i=1}^N T_i y_i(1)}{n_1} \right] =
\frac{1}{n_1} \sum_{i=1}^N E(T_i) y_i(1)  

\end{equation}

<br/>

* The (true) Average Treatment Effect is given by:

<p><p/>
\begin{equation}
\begin{aligned}
\mathrm{ATE} & =y \overline{(1)}-y \overline{(0)} \\
\overline{y(1)} & =\frac{\sum_{i=1}^N y_i(1)}{N} \\
\overline{y(0)} & =\frac{\sum_{i=1}^N y_i(0)}{N}
\end{aligned}
\end{equation}

* The (true) Average Treatment Effect is given by:

\begin{equation}
\text{ATE} = \bar{y(1)} - \bar{y(0)} \\
\bar{y(1)} = \frac{\sum_{i=1}^N y_i(1)}{N} \\
\bar{y(0)} = \frac{\sum_{i=1}^N y_i(0)}{N} \\
\label{eq:ATE-def}
\end{equation}
* The difference-in-means estimator of ATE:

\begin{equation}
\widehat{\text{ATE}} = \widehat{\bar{y(1)}} - \widehat{\bar{y(0)}} \\
\widehat{\bar{y(1)}} = \frac{\sum_{i=1}^N T_i y_i(1)}{n_1} \\
\widehat{\bar{y(0)}} = \frac{\sum_{i=1}^N C_i y_i(0)}{n_0} \\
\label{eq:DME-ATE-def}
\end{equation}



<!-- To run
# nohup Rscript -e "knitr::knit2html('_M5A-Neyman_1923.Rmd')" > _M5A-Neyman_1923.log  &

# Or
# nohup Rscript -e "rmarkdown::render('_M5A-Neyman_1923.Rmd')" > _M5A-Neyman_1923.log  &


-->
