---
title: Professional Development Plan for Experienced Statisticians
subtitle: Bridging Traditional Methods with Modern Approaches
output:
  # for use of .tabset
  bookdown::html_document2:
    code_download: true
    code_folding: hide
    toc: false
    fig_caption: yes
    number_sections: yes
    css: ../../css/pandoc3.css
bibliography: [../_bibFiles/_healthy_aging.bib, ../_bibFiles/_ref_free_genom_inf.bib, ../../_bibFiles/_Breiman.bib, ../../_bibFiles/_Yu.bib, ../../_bibFiles/_Freedman.bib, ../../_bibFiles/_RUV.bib, ../../_bibFiles/_Yu.bib, ../../_bibFiles/_rand_cause_inf.bib]
csl: ../_csl/cell-numeric.csl
link-citations: true
---

 

<style>

.watermark {
  opacity: 0.2;
  position: fixed;
  top: 50%;
  left: 50%;
  font-size: 500%;
  color: #00407d;
}

</style>

---


```{r GlobalOptions, results="hide", include=F, cache=FALSE}

knitr::opts_knit$set(stop_on_error = 2L) #really make it stop
options(knitr.table.format = 'html')

```

```{r prof-dev-plan-Prelims, include=F, echo=FALSE, results='hide', message=FALSE,cache=F}

FN <- "_prof_dev_plan"
if(sum(grepl(FN, list.files()))==0) stop("Check FN")

PREFIX <- "_prof_dev_plan" #- replace by FLOWCELL???

suppressPackageStartupMessages(require(methods))  
suppressPackageStartupMessages(require(rmarkdown))  
suppressPackageStartupMessages(require(bookdown))  

suppressPackageStartupMessages(require(knitr))  
options(stringsAsFactors=F)  

suppressPackageStartupMessages(require(data.table)) 
options(datatable.fread.datatable=F)

suppressPackageStartupMessages(require(plyr))
suppressPackageStartupMessages(require(dplyr))
suppressPackageStartupMessages(require(magrittr))

# Shotcuts for knitting and redering while in R session (Invoke interactive R from R/Scripts folder)
kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
      output=paste(FN,".html", sep=''))

rr <- function(n='') rmarkdown::render(paste("t", n, sep=''), envir=globalenv(),
      output_file=paste(FN,".html", sep='')) ##, output_dir='Scripts')

bb <- function(n='') browseURL(paste(FN,".html", sep=''))

# The usual shotcuts
zz <- function(n='') source(paste("t", n, sep=''))

# Using relative paths:
# Script will be invoked from R/Scripts folder.  Make 'R' the WRKDIR.
WRKDIR <- ('..')

# Not needed if path is relative ...
if(!file.exists(WRKDIR)) stop("WRKDIR ERROR: ", WRKDIR)

# do once
#setwd(WRKDIR)

# file rmarkdown file management options: cache, figures
cache_DIR <- file.path(WRKDIR, 'Scripts', 'cache/index/')
suppressMessages(dir.create(cache_DIR, recursive=T))
opts_chunk$set(cache.path=cache_DIR)

# NOTE: need to add FC to figures path
# due to confluence "feature" of keeping the filename of embedded figures

figures_DIR <- file.path(WRKDIR, 'Scripts', 'figures/index/')
#figures_DIR <- file.path('figures/index')

suppressMessages(dir.create(figures_DIR, recursive=T))
opts_chunk$set(fig.path=paste0(figures_DIR))###, FC))

tables_DIR <- file.path(WRKDIR, 'Scripts', 'tables/index/')
suppressMessages(dir.create(tables_DIR, recursive=T))

# need a local copy of help_DIR
#help_DIR <- file.path(WRKDIR, 'Scripts', 'help_files')
help_DIR <- file.path('.', 'help_files')
suppressMessages(dir.create(help_DIR, recursive=T))

#temp_DIR <- file.path(WRKDIR, 'Scripts', 'temp_files')
#suppressMessages(dir.create(temp_DIR, recursive=T))


```
<!-- ######################################################################## -->


```{r index-utilityFns, echo=T, include = F}

 source('utilityFns.r')
 #source('utilities/Mymva.r')

round_df <- function(x, digits) {
    # round all numeric variables
    # x: data frame 
    # digits: number of digits to round
    numeric_columns <- sapply(x, mode) == 'numeric'
    x[numeric_columns] <-  round(x[numeric_columns], digits)
    x
}
```
<!-- ######################################################################## -->

This professional development plan is designed for statisticians with 10-15+ 
years of experience who need to update their skills with the latest developments
in:
* statistical computing
* analysis
* reporting methods

The professional development program 
was created with assistance from Claude, an AI assistant developed by Anthropic.
Claude helped structure the training modules, generate activity descriptions,
and organize the implementation timeline based on current trends in statistical
computing and modern methodologies. The AI incorporated feedback and
made targeted revisions to sections about machine learning methodologies
and computing environments, demonstrating how AI tools can be used
collaboratively for curriculum development.

<br/>

## Learning Track 1: Modern Statistical Computing Environments 

### Transition from Traditional Software to Modern Platforms
**Format**: Hands-on workshop (3 hours)  

**Objective**: Familiarize with modern statistical computing environments that complement or replace SAS, SPSS, or older R versions

**Activities**: 

- Comparative exercise: Solve the same analysis problem in traditional software and then in Python/modern R
- Pair programming: Team up experienced statisticians with those already familiar with modern tools
- Migration exercise: Convert an existing analysis script from legacy software to modern alternatives

### Version Control and Reproducible Analysis
**Format**: Interactive tutorial with practice (2 hours)

**Objective**: Master Git/GitHub for code versioning and collaboration

**Activities**: 

- Set up personal GitHub repositories for statistical projects
- Practice creating branches, commits, and pull requests
- Implement a reproducible analysis workflow using Git, R Markdown or Jupyter Notebooks
- Review each other's repositories and provide feedback on structure and documentation

### Cloud Computing for Statistical Analysis
**Format**: Guided exploration (2 hours)   

**Objective**: Utilize cloud platforms for scalable statistical analysis

**Activities**:  

- Set up environments on platforms like AWS, Google Cloud, or Azure
- Deploy and run statistical models on cloud infrastructure
- Compare performance between local and cloud-based computation for large datasets
- Collaborative challenge: Develop a cloud-based analysis pipeline for a complex dataset

<br/>

## Learning Track 2: Modern Statistical Methods and Approaches

### AI and LLM Integration for Statistical Work
**Format**: Workshop with hands-on exercises (4 hours)

**Objective**: Leverage AI assistants and large language models to enhance statistical workflows

**Activities**:  

- Explore effective prompting techniques for statistical analysis with models like GPT-4 and Claude
- Practice using AI assistants for code debugging, explanation, and optimization
- Develop custom workflows that combine human statistical expertise with AI capabilities
- Critical evaluation: Assess the limitations and potential biases when using AI for statistical work
- Ethical considerations workshop: Discuss responsible AI use in professional statistical practice

- Integration exercise: Design a hybrid workflow that leverages both traditional statistical methods and AI tools

### Bayesian Methods in Practice
**Format**: Case-study workshop (4 hours)   

**Objective**: Apply Bayesian approaches to real-world statistical problems

**Activities**:  

- Compare frequentist vs. Bayesian approaches to the same problem
- Implement Bayesian models using modern tools (Stan, PyMC3, brms)
- Develop intuition for prior selection through interactive simulations
- Group challenge: Design and implement a Bayesian solution to an industry-relevant problem

### Machine Learning Integration
**Format**: Project-based learning (6 hours split across multiple sessions)   

**Objective**: Effectively integrate machine learning with traditional statistical approaches

**Activities**:  

- Identify appropriate use cases for ML vs. traditional statistical methods
- Develop a taxonomy of problems and corresponding methodological approaches
- Implement hybrid models that combine statistical rigor with ML flexibility
- Critical evaluation exercise: Analyze published papers that use ML methods for statistical inference

### Causal Inference Methods
**Format**: Seminar with practical exercises (4 hours)   

**Objective**: Master modern causal inference frameworks and methods

**Activities**:  

- Compare traditional regression approaches with modern causal inference methods
- Implement causal graphs and identify assumptions using DAGitty or similar tools
- Apply methods like propensity score matching, instrumental variables, and difference-in-differences
- Design challenge: Develop a causal inference strategy for a complex observational dataset

<br/>

## Learning Track 3: Modern Reporting and Communication  {-}

### Dynamic Reporting and Dashboards
**Format**: Hands-on workshop (3 hours)   

**Objective**: Create interactive statistical reports and dashboards

**Activities**:  

- Convert static reports to interactive documents using Shiny, Dash, or Observable
- Develop parameterized reports that stakeholders can customize
- Implement effective data visualization principles in interactive contexts
- Peer review: Evaluate and provide feedback on each other's interactive reports

### Communicating Uncertainty
**Format**: Practice session with feedback (2 hours)
**Objective**: Effectively communicate statistical uncertainty to non-technical stakeholders

**Activities**:  

- Develop visual representations of uncertainty beyond error bars and p-values
- Create explanatory tools for Bayesian credible intervals and posterior distributions
- Role-play exercise: Explain complex statistical results to simulated stakeholders
- Design challenge: Create a one-page uncertainty summary for a complex analysis

### Automated Reporting Pipelines
**Format**: Project-based learning (4 hours)   

**Objective**: Implement automated statistical reporting workflows

**Activities**:  

- Set up GitHub Actions or similar CI/CD tools for automated report generation
- Create templated analyses that can be applied to regularly updated data
- Develop quality control checks that run automatically
- Team challenge: Design and implement an end-to-end automated analysis pipeline

<br/>
## Implementation Plan {-}

### Month 1: Foundations
- Week 1-2: Modern Computing Environments (Activities 1.1 and 1.2)
- Week 3-4: AI Integration and Introduction to Modern Methods (Activities 2.1 and 2.2)

### Month 2: Advanced Methods
- Week 5-6: Machine Learning Integration (Activity 2.3)
- Week 7-8: Causal Inference (Activity 2.4)

### Month 3: Communication and Application
- Week 9-10: Modern Reporting (Activities 3.1 and 3.2)
- Week 11-12: Automated Pipelines and Integration (Activity 3.3)
- Final Week: Capstone Project

<br/>

## Assessment Strategy

### Formative Assessment
- Regular peer feedback sessions
- Self-assessment reflection logs
- Incremental project milestones

### Summative Assessment
- Capstone project: Participants apply all learned skills to a real-world statistical problem
- Portfolio of work demonstrating proficiency in new methods and tools
- Presentation to peers on a chosen advanced topic

<br/>

## Resource Requirements

### Technology
- Access to cloud computing resources (AWS/GCP/Azure credits)
- Modern statistical software installations (R/RStudio, Python/Jupyter)
- Version control systems (Git)
- Interactive reporting tools (R Shiny, Dash, Observable)

### Personnel
- Lead instructor with expertise in both traditional and modern statistical methods
- Topic specialists for advanced sessions
- Technical support for software installation and troubleshooting

### Materials
- Curated datasets representing diverse analytical challenges
- Pre-configured computing environments to minimize setup time
- Reference guides for transitioning between software platforms

<br/>
1
## Extension Activities

### Community of Practice
- Establish a peer learning community for ongoing support
- Regular journal club focused on emerging methods
- Mentorship pairing between participants with complementary strengths

### Continuous Learning Paths
- Specialized deep-dive workshops on specific tools or methods
- Advanced certification opportunities
- Research collaboration opportunities

---

This professional development plan is designed to be modular, allowing participants to focus on areas most relevant to their specific needs while ensuring comprehensive coverage of modern statistical practices.

```{r, echo=FALSE}
  knitr::knit_exit()
```


<!-- To run
# nohup Rscript -e "knitr::knit2html('_prof_dev_plan.Rmd')" > _prof_dev_plan.log  &
    
# Or
# nohup Rscript -e "rmarkdown::render('_prof_dev_plan.Rmd')" > _prof_dev_plan.log  &

-->


