---
title: Statistics Training and Data Re-Analysis for Better Results
#subtitle: "a tale of two schools"
author: "[Francois Collin](https://www.linkedin.com/in/francoisz/)"
date: "`r format(Sys.time(), '%B %d, %Y')`"
always_allow_html: yes
output:
  # for gtanse of .tabset
  bookdown::html_document2:
  #html_document:
    code_folding: hide
    code_download: true
    toc: true
    toc_depth: 3
    # does this have an effect
    fig_caption: yes
    # this has no effect
    number_sections: yes
    # css: ['../_css/pandoc3.css', '../_css/myMargins.css']
bibliography: [../_bibFiles/_healthy_aging.bib, ../../_bibFiles/_Breiman.bib, ../../_bibFiles/_Freedman.bib, ../../_bibFiles/_Yu.bib, ../../_bibFiles/_RUV.bib, ../../_bibFiles/_RMA.bib]
csl: ../_csl/cell-numeric.csl
link-citations: true
---

```{css sidenote, echo = FALSE}

.main-container {
    margin-left: 250px;
}
.sidenote, .marginnote { 
  float: right;
  clear: right;
  margin-right: -40%;
  width: 37%;         # best between 50% and 60%
  margin-top: 0;
  margin-bottom: 0;
  font-size: 1.1rem;
  line-height: 1.3;
  vertical-align: baseline;
  position: relative;
  }
```


<style>
@import url('https://fonts.googleapis.com/css?family=Raleway');
@import url('https://fonts.googleapis.com/css?family=Oxygen');
@import url('https://fonts.googleapis.com/css?family=Raleway:bold');
@import url('https://fonts.googleapis.com/css?family=Oxygen:bold');

.main-container {
  max-width: 1400px !important;
}

body{
  font-family: 'Oxygen', sans-serif;
  font-size: 16px;
  line-height: 24px;
}

h1,h2,h3,h4 {
  font-family: 'Raleway', sans-serif;
}

.container { width: 1400px; }

caption {
  font-size: 20px;
  caption-side: top;
  text-indent: 30px;
  background-color: lightgrey;
  color: black;
  margin-top: 5px;
}

g-table-intro h4 {
  text-indent: 0px;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      fig.width = 8,
                      fig.height = 4)
```

```{r infex-GlobalOptions, results="hide", include=FALSE, cache=FALSE}

knitr::opts_knit$set(stop_on_error = 2L) #really make it stop
options(knitr.table.format = 'html')

options(stringsAsFactors=F)

 #knitr::dep_auto()

```
<!-- ######################################################################## -->


```{r infex-Prelims,  include=FALSE, echo=FALSE, results='hide', message=FALSE} 


FN <- "index"
if(sum(grepl(FN, list.files()))==0) stop("Check FN")

PREFIX <- "index" #- replace by FLOWCELL???

 suppressMessages(require(rmarkdown))
 suppressMessages(require(knitr))

 suppressPackageStartupMessages(require(methods))
 suppressPackageStartupMessages(require(bookdown))

 suppressPackageStartupMessages(require(data.table))
 options(datatable.fread.datatable=F)

 suppressPackageStartupMessages(require(plyr))
 suppressPackageStartupMessages(require(dplyr))
 suppressPackageStartupMessages(require(magrittr))

 # Shotcuts for knitting and rendering while in R session (Invoke interactive R from R/Scripts folder)
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste(FN,".html", sep=''))

 rr <- function(n='') rmarkdown::render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep='')) ##, output_dir='Scripts')

 bb <- function(n='') browseURL(paste(FN,".html", sep=''))

 # The usual shortcuts
 zz <- function(n='') source(paste("t", n, sep=''))


 WRKDIR <- '..'
 if(!file.exists(WRKDIR)) stop("WRKDIR ERROR", WRKDIR)

 # do once

 # Shotcuts for knitting and rendering while in R session
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste('',FN,".html", sep=''))

 rr <- function(n='') render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep=''), output_dir='Scripts')

 bb <- function(n='') browseURL(paste('',FN,".html", sep=''))

 # The usual shorcuts
 zz <- function(n='') source(paste('', "t", n, sep=''))

 # file rmarkdown file management options: cache, figures
 cache_DIR <- file.path(WRKDIR,'Scripts', 'cache/M1A/')
 suppressMessages(dir.create(cache_DIR, recursive=T))
 opts_chunk$set(cache.path=cache_DIR)

 figures_DIR <- file.path(WRKDIR,'Scripts', 'figures/M1A/')
 suppressMessages(dir.create(figures_DIR, recursive=T))
 opts_chunk$set(fig.path=figures_DIR)

 #tables_DIR <- file.path(WRKDIR,'Scripts', 'tables/M1A/')
 #suppressMessages(dir.create(table_DIR, recursive=T))
 #opts_chunk$set(fig.path=table_DIR)
 
 # need a local copy of help_DIR
 #help_DIR <- file.path(WRKDIR,'Scripts', 'help_files')
 help_DIR <- file.path('.', 'help_files')
 suppressMessages(dir.create(help_DIR, recursive=T))
 
 temp_DIR <- file.path(WRKDIR,'Scripts', 'temp_files')
 suppressMessages(dir.create(temp_DIR, recursive=T))

```
<!-- ######################################################################## -->

*** 

```{r infex-utilityFns, echo=FALSE}
 # Here we define some utility functions
source('r/utilityFns.r')

```

<!-- ######################################################################## -->

```{r infex-input-params-all, echo = FALSE, results = "asis", eval=F, echo=F}
 # print out original input params
data.frame(
param_name=names(params),
param_value=unlist(params),
param_class=sapply(params, class), row.names=NULL) %>%
  knitr::kable(caption="Input Parameters")
```

<br/>


<!--
* Breiman: [@Breiman:1984aa; @Breiman:2001aa]
* Freedman: [@Freedman:2008aa; @Freedman:2008ad; @Freedman:2008ac; @Freedman:2009ac]
-->



# Challenges in Statistical Analysis of Genomic Data {-}


Results of statistical data analysis can be improved for various reasons:
<p><p/>
* the analysis relies on **inappropriate statistical models** - Freedman: 
[@Freedman:2008aa; @Freedman:2008ad; @Freedman:2008ac; @Freedman:2009ac]
   - Professor Freedman focused on the use of models in the social sciences
   - Inappropriate modeling is not limited to the social sciences
      - in biotech, the unchecked use of logistic regression whenever the
response is binary is ubiquitous.
         - the results of such analyses could be misleading due to the biases
illustrated in Freedman (2008) [@Freedman:2008aa]

<p><p/>
* the analysis is flawed due to the reliance on **standard textbook methods** which
have a limited range of applicability - Breiman:
[@Breiman:1984aa; @Breiman:2001aa]
   - as regression models are used to answer almost every question
that come up in the internal analyses conducted by companies in industry,
many are going to be flawed.  
      - For a review of how regression models go wrong see
[@Freedman:2009aa;@Freedman:2010aa;@Freedman:2008ac;@Freedman:2008ad;@Freedman:2004aa;@Freedman:2008aa;@Freedman:2008ab]
   - clinical validation studies for complex omic Dx instruments that are designed 
according to FDA guidelines will make use of the classical Neyman-Pearson hypothesis
testing context which is often invalid:
      - the samples in the study may not be representative of the target population.
         - there is rarely any explicit randomness in the selection of the study samples
so inference to any superset of the study samples is iffy at best.
      - dependency structure among samples processed in batches is often complex
         - a run may be treated as a batch with internal structure or sub-batches.
      - N is often small 
      - contrived samples may not be interpretable as intended or required
      - other???

<p><p/>
* many analyses may be improved with the help of
**valid assessments of performance** - Speed:
[@Irizarry:2003aa; @Bolstad:2004aa;@Gagnon-Bartsch:2012aa;@Gandolfo:2018aa;@Irizarry:2006aa;@Jacob:2018aa;@Risso:2014aa;@Yuan:2019aa]
   - *valid* here means that the assessment provides a direct indicator of performance 
on a meaningful scale or end point.
      - when normalizing RNASeq gene expression profiles, one can measure the effectiveness of
the normalization step by examining the similarities among gene expression profiles before
and after normalization.  Although this is a sensible way to assess the effect of the 
normalization step, it is not a good metric to assess the impact normalization has on
the ultimate goal of the analysis - DGE analysis.
      - when comparing two sequencing instruments in terms of their ability to carry-out
a reagent QC protocol designed to assess reagent quality:
         - although the appropriate assessment probably calls for an assessment of
performance against the truth, and a comparison of these performance parameters between
instruments, regulatory compliance may call for an equivalency assessment
            - it should not take much reflecting realize that an equivalency assessment
creates unnecessary complications and can go wrong in many ways:
               -  an obvious problem being drift.
               - down sampling to make the two instruments comparable is another obvious
problem
      - using an upstream measurement to predict unconverted fraction
         -SCALE, SCALE, SCALE



<br/>

# Statistics Training {-}


## Design Training Program  with AI {-}

As an exercise working witch claude.ai, we design a training program for experienced statisticians.
The training is meant to bring statisticians up to speed with novel statistical methods and techniques.

* [stats-pd-plan-final-v35.html](./_M1C-stats-pd-plan-final-v35.html) - final version of the
professional development program.

* [stats-pd-plan-change-log.html](./_M1C-stats-pd-plan-change-log.html)  -
detailed change log documenting the development process

* [ai-assistant-guide](./_M1C-ai-assistant-guide.html) - outline
principles which apply broadly to working with AI assistants on complex projects.

 
<br/>

## Training Resources  {-}

* Gen AI

* Deep Learning Models


<br/>

# Beyond RUV {-}

Can the methodology which enables the basic RUV analysis be adapted to
other applications?

<p><p/>
* IUV - identification of unwanted variability
   - the automated high throughput processing of samples necessary
to get the omic based readouts used to diagnose samples is carried out by
a pipeline which strings together numerous steps (typically hundreds).
      - eg. Cell-Free DNA Methylation Profiling Analysis
   - As samples proceed through the pipeline they will be organized in various configurations
giving rise to various groupings:
      - plasma isolation
      - cfDNA extraction 
      - amplification, conversion, ...
   - measurements are recorded as samples travel through the pipeline, each
measurement potentially having its own shared variability or dependency structure 
   - a real problem is to identify the factors giving rise to excessive variability
in the downstream read-outs


<p><p/>
* QUV - quantification of unwanted variability
   - can RUV methodology be adapted to provide quantification of reproducibility?

<br/>

<p><p/>
# Data (Re-)Analysis Projects {-}

## Reference-Free Genomic Inference with SPLASH and OASIS {-}

* [_M2A-RFGI_v1.html](./_M2A-RFGI_v1.html) -
* [_M2B-RFGI_v2.html](./_M2B-RFGI_v2.html) -
* [_M3A-OASIS.html](./_M3A-OASIS.html) -

<br/>

<p><p/>
## Miscellanious  {-}
  - [_M4A-intro_IUV.html](./_M4A-intro_IUV.html) 
  - [_M5A-cf_ddPCR_seq.html]
  - [_M6A-randomness_in_calls.html] 
  - [_M7A-interimSampleSizeAnalysis.html]
      - [_M3A_study_design_CB_notes.pdf]


<br/>

# Statistics Training  


## [ 

***


<br/>

# References
<div id="refs"></div>


<br/>

# Appendix: Full list of reports 

```{r , child=c('_index.Rmd')}
```


<br/>




```{r, echo=FALSE}
  knitr::knit_exit()
```


