---
title: "Statistical Analysis of Data in the Biotech Industry"
###\nLessons from 20 years of expererience"
author: "[Francois Collin](https://www.linkedin.com/in/francoisz/)"
date: "`r format(Sys.time(), '%B %d, %Y')`"
always_allow_html: yes
output:
  #html_document:
  # for use of .tabset
  bookdown::html_document2:
    code_folding: hide
    code_download: true
    toc: true
    toc_depth: 3
    # does this have an effect
    fig_caption: yes
    # this has no effect
    number_sections: yes
    # css: ['../_css/pandoc3.css', '../_css/myMargins.css']
bibliography: [../_bibFiles/_healthy_aging.bib, ../../_bibFiles/_Breiman.bib, ../../_bibFiles/_Freedman.bib, ../../_bibFiles/_Yu.bib, ../../_bibFiles/_RUV.bib, ../../_bibFiles/_RMA.bib, ../../_bibFiles/_scRNAseq_norm.bib]
csl: ../_csl/cell-numeric.csl
link-citations: true
---

```{css sidenote, echo = FALSE}

.main-container {
    margin-left: 250px;
}
.sidenote, .marginnote { 
  float: right;
  clear: right;
  margin-right: -40%;
  width: 37%;         # best between 50% and 60%
  margin-top: 0;
  margin-bottom: 0;
  font-size: 1.1rem;
  line-height: 1.3;
  vertical-align: baseline;
  position: relative;
  }
```


<style>
@import url('https://fonts.googleapis.com/css?family=Raleway');
@import url('https://fonts.googleapis.com/css?family=Oxygen');
@import url('https://fonts.googleapis.com/css?family=Raleway:bold');
@import url('https://fonts.googleapis.com/css?family=Oxygen:bold');

.main-container {
  max-width: 1400px !important;
}

body{
  font-family: 'Oxygen', sans-serif;
  font-size: 16px;
  line-height: 24px;
}

h1,h2,h3,h4 {
  font-family: 'Raleway', sans-serif;
}

.container { width: 1400px; }

caption {
  font-size: 20px;
  caption-side: top;
  text-indent: 30px;
  background-color: lightgrey;
  color: black;
  margin-top: 5px;
}

g-table-intro h4 {
  text-indent: 0px;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE,
                      cache = FALSE,
                      fig.width = 8,
                      fig.height = 4)
```

```{r infex-GlobalOptions, results="hide", include=FALSE, cache=FALSE}

knitr::opts_knit$set(stop_on_error = 2L) #really make it stop
options(knitr.table.format = 'html')

options(stringsAsFactors=F)

 #knitr::dep_auto()

```
<!-- ######################################################################## -->


```{r index-Prelims,  include=FALSE, echo=FALSE, results='hide', message=FALSE} 


FN <- "index"
if(sum(grepl(FN, list.files()))==0) stop("Check FN")

PREFIX <- "index" #- replace by FLOWCELL???

 suppressMessages(require(rmarkdown))
 suppressMessages(require(knitr))

 suppressPackageStartupMessages(require(methods))
 suppressPackageStartupMessages(require(bookdown))

 suppressPackageStartupMessages(require(magrittr))

 # Shotcuts for knitting and rendering while in R session (Invoke interactive R from R/Scripts folder)
 kk <- function(n='') knitr::knit2html(paste("t", n, sep=''), envir=globalenv(),
       output=paste(FN,".html", sep=''))

 rr <- function(n='') rmarkdown::render(paste("t", n, sep=''), envir=globalenv(),
       output_file=paste(FN,".html", sep='')) ##, output_dir='Scripts')

 bb <- function(n='') browseURL(paste(FN,".html", sep=''))

 # The usual shortcuts
 zz <- function(n='') source(paste("t", n, sep=''))


```
<!-- ######################################################################## -->

<br/>


<!--
* Breiman: [@Breiman:1984aa; @Breiman:2001aa]
* Freedman: [@Freedman:2008aa; @Freedman:2008ad; @Freedman:2008ac; @Freedman:2009ac]
-->


# Reference-Free Genomic Inference with SPLASH and OASIS

## Introduction


<p><p/>
* The [Salzman Lab](https://salzmanlab.stanford.edu/) 
develop and apply new **statistical algorithms** for **biological inference**
on fundamental questions in the evolution of genomes.

<p><p/> 
* Here we consider 2 papers from the [Salzman Lab](https://salzmanlab.stanford.edu/):
   - SPLASH: 
A statistical, reference-free genomic algorithm unifies biological discover
[PDF](https://www.cell.com/cell/pdf/S0092-8674(23)01179-0.pdf),
[Code](https://github.com/refresh-bio/SPLASH) [@Chaung:2023aa]
   - OASIS: 
An interpretable, finite-sample valid alternative to Pearson’s X2 
for scientific discovery
[URL](https://www.pnas.org/doi/10.1073/pnas.2304671121),
[Code](https://github.com/refresh-bio/SPLASH) [@Baharav:2024aa]


The articles provide a proof of concept of the proposed methods.

Primary Criticism:

* Technical validation ion the omic scale is required to get a sense of the utility of the
proposed methods.  

* **Technical validation** should be run on typical data and should include both
measures of algorithm performance and resistance to **input data quality** and
**unwanted variation**.  See [intro_2_UV](...).

* The claim of the validity of the p-values, one of the selling points of the method,
is suspect.  When we fit one prototype 
model to 20K, 50K or 100K feature data subsets we do not believe that the  model
fits adequately in every case: we may robustify the fit to allow for the 
presence of bad points, or outliers, and we do not put too much credibility on the
precise values of p-values; we believe that they generally order hypotheses the
right way.
   - In this particular case there are bigger problems with the p-values -
their properties are derived based in the assumption that $$f$$ and $$c$$ 
are fixed values, when they are derived from the data.


### On End to End Omic Scale Analysis 

<p><p/>
* Omic scale data refers to measurements taken in parallel over many features -
from thousands of genes to millions of proteins.  Even though the result of
interest, gene scores for example, may be obtained from analyzing
features one at a time, the success of the design and analysis methods
can only be assessed by taking a holistic look at the study data and results.

* To have a holistic picture, the assessment must start with the input material
and follow the sample set through all of the processing steps including QC assessment
and filtering (if any), sample pre-processing  (further filtering and normalization),
data analysis (compute the desired results and report the findings).

* The article provided the information needed to
show that the technology works in principle, but to appreciate
how well the technology will work in practice, it has to be evaluated 
in a way that fully reflects the  expected usage.

<!--
* Having good comparator analysis methods is helpful, if not crucial.  I'm not sure that
the chi-square test provides a good comparator.  I don't recall ever coming 
across the chi-square test in the course of RNA-Seq related analyses, but
my experience is limited to comparing gene expression levels across samples.
There are a number of R packages which treat compositional data analysis
which is analogous to the problem addressed in SPLASH.  Many options 
are listed in Section \@ref(rna-seq-wi-samp); it should be possible to customize one or more
to provide a comparator for SPLASH+OASIS.
-->

* Having a good dataset for assessment
is also critical and could include control features, control samples,
and samples providing engineered or designed truth.
In addition to having built-in truth, the samples used to assess
performance must also be representative of the actual samples
which will be processed in production.

<br/>

<!--
## On inferential frameworks   {-}

* An important feature of the SPLASH+OASIS data analyses pipeline is the
finite sample valid statistical inference that it makes possible whereas other
methods rely on asymptotically valid estimates. The value that this feature confers to 
the method should be verified empirically with a proper omic level assessment.  Note that RNA-Seq
methods rely heavily in these asymptotically valid estimates which have not been found
to be problematic for small sample sizes  as the site-specific estimates are typically
stabilized by shrinkage.


* From a personal perspective I feel that the write-up of the papers puts a lot of stock
in p- and q-values.  The statistical standard errors of measurement, or the sampling errors,
are only one source of variability in data collected in a high throughput, massively
parallel fashion. Other sources of variability such as reagent lots, instruments,
operators, days (ie. day to day variability is variability that is not otherwise specified)
will often dominate the total variability.  

* In practice we select the analysis method which
produces the best results as ascertained by omic scale assessments of accuracy.
-->



## Presentation

* The presentation [linked here](_M3B-SPLASH-OASIS/splash_oasis_methodology_presentation_v2.html)
provides a high level summary of the SPLASH article in the words of the authors. 

* Some comments are inserted where there might be gaps or a difference of opinion.

<br/>

## [SPLASH](_M2B-RFGI_SPLASH_v2.html)

* Contains a literature review  of Reference-Free Genomic Inference prior to SPLASH, as well
methods developed for RNA-seq which could be used here.



## [OASIS](./_M3A-OASIS.html) 


* Contains some notes on OASIS ,.


<br/>

# The Identification and Removal of Unwanted Variability

* RUV might be the most imoortat missibf oiece for SPLASH

<br/>

# References
<div id="refs"></div>

<br/>

# Appendix: 



```{r, echo=FALSE}
  knitr::knit_exit()
```

################################################
## ARCHIVE
################################################

